---
title: "Reproducable Research Peer Assessment 1"
output: html_document
---



## Loading and preprocessing the data
Read data file into a dataframe and inspect the data.
```{r}
data = read.csv('activity.csv', header = T)
names(data)
head(data)
dim(data)
```

Transform the date field into the date class 
```{r}
date <- strptime(data$date, "%Y-%m-%d")
data$date <- date
```

Get unique dates for use in subsequent calculations
```{r}
uniqueDates <- unique(data$date)
```

Get unique intervals for use in subsequent calculations
```{r}
uniqueIntervals <- unique(data$interval)
```


## What is mean total number of steps taken per day?
Create a histogram of number of steps taken per day.
```{r}
# split and group data by the day of the year (yday).
# Example: 1st Jan => yday = 1, 31st Dec => yday = 365
groupByDays <- split(data$steps, data$date$yday)

# Find the sum of steps each day of the year, ignoring empty values
stepsEachDay <- sapply(groupByDays, sum, na.rm=TRUE)

# Plot histogram with x-axis as day of the year
# and the y-axis as the total number of steps 
plot(uniqueDates, stepsEachDay, main="Histogram of Steps Each Day", 
     xlab="Date", ylab="Frequency", type="h", lwd=4, col="blue")

```

Calculating the mean of number of steps each day
```{r}
meanStepsEachDay <- sapply(groupByDays, mean, na.rm=TRUE)
dfMeanStepsEachDay <- data.frame(date=uniqueDates, mean=meanStepsEachDay, row.names=NULL)
dfMeanStepsEachDay

```

Calculating the median of number of steps each day
```{r}
medianStepsEachDay <- sapply(groupByDays, median, na.rm=TRUE)
dfMedianStepsEachDay <- data.frame(date=uniqueDates, median=meanStepsEachDay, row.names=NULL)
dfMedianStepsEachDay

```

## What is the average daily activity pattern?

```{r}
# split and group data by the interval field.
groupByIntervals <- split(data$steps, data$interval)

# Find the mean of steps for each interval, ignoring empty values
meanStepsEachInterval <- sapply(groupByIntervals, mean, na.rm=TRUE)

# Plot the time-series
plot(uniqueIntervals, meanStepsEachInterval, type="l",
     main="Average Number of Steps per Interval", 
     xlab="Interval", ylab="Average # of steps", 
     lwd=2, col="blue")

# Find the location of where the maximum is
maxMeanStepsEachInterval <- max(meanStepsEachInterval, na.rm=TRUE)
maxMeanStepsEachIntervalIdx <- which(meanStepsEachInterval == maxMeanStepsEachInterval)
maxInterval <- uniqueIntervals[maxMeanStepsEachIntervalIdx]

# Print the location of the maximum interval
maxInterval

```


## Inputing missing values

Calculate total amount of missing values in the data set:
```{r}
booleanComplete <- complete.cases(data$steps)
sumMissing <- sum(as.numeric(!booleanComplete))
sumMissing

```


